### 2017-07-22 chapters/introduction/physiology.tex
Add latest

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.7\textwidth]{chapters/introduction/figures/heart_anatomy.pdf}
\caption{The heart anatomy. (Adapted and modified from Wapcaplet in Sodipodi).}
\label{fig:heart_anatomy}
\end{figure}

Deoxgenated blood is transported from the body trough the veins and to
the right atrium (RA). From the RA, blood flows to the right ventricle
(RV), before it is sent through the the pulmonary arteries(PA) and to the
lungs, where the bloods gets oxygenated. From the lungs, the blood
travels trough the pulmonary veins to the left atrium (LA), and
finally to the left ventricle (LV), before is is sent out to the body
again trough the aorta.

@@

Starting from the outside of the heart we have the \emph{pericardium},
the \emph{epicardium}, the \emph{myocardium} and the \emph{endocardium}. 
The endo- and epicardium are respectively the inner and outer layers of the
myocardium. The myocardium is made up by individual muscle cells, or
myocytes, which branch and join neighboring cells, and form a
complicated fibrous network which is often referred to as a 
\emph{functional syncytium} (meaning that the number of active muscle
fibers cannot be varied from beat to beat). Each myocyte is composed
of bundles of myofibriles, which again contains
\emph{myofilaments}. The myofilaments consists of a repeated pattern
of lines and bands which can be seen as a collection of individual
contractile units called \emph{sarcomeres}. A simple sketch of a
sarcomere is shown in Figure \ref{fig:sarcomere}. 
\begin{figure}[htbp]
  \centering
    \includegraphics[width=1.0\textwidth]{chapters/introduction/figures/Sarcomere.png}
\caption{Illustration of the sarcomere structure.}
\label{fig:sarcomere}
\end{figure}

The sarcomeres are composed thick and thin filaments that slides
back and forth when the muscle contract or relaxes.
The thin filaments are made up by a protein called \emph{actin} while
the thick filaments are made up by a protein called \emph{myosin}.

@@

The heart beats in a cyclic manner, about one beat every second.
The contraction of the heart is triggered by an electrical signal called an action
potential originating from specialized cells in the right atrium. When an action
potential is triggered, calcium is released into the cells. These
calcium ions binds to a complex called troponin C located at the thin
filaments, which then exposes the actin to the myosin head. When
myosin binds to actin we say that a \emph{cross-bridge} is formed
between the thin and thick filaments. A \emph{power stroke} is
triggered after ATP hydrolysis, and the sarcomere shortens. 
The signal propagates through the myocardium but also along
specialized electrical highways with high conducting cells.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.85\textwidth]{chapters/introduction/figures/cardiac_cycle.png}
\caption{Showing the relation between pressure and volume in the left
  ventricle during the cardiac cycle, which the different phases and
  valvular events marked. To the left we show the a reduced version of
the classical Wiggers diagram with pressure and volume plotted against
time on the $x-$axis. To the right we show the pressure volume loop
with volume plotted on the $x-$axis and pressure on the $y-$axis. }
\label{fig:pv_loop}
\end{figure}

@@

The pressure and volume inside the ventricle can provide us with
information about the passive and active properties of the
myocardium. Imaging that it was possible to turn of all contractile
units in the myocardium, and remove all loads (e.g set the pressure
to zero). This would be analogous to have a deflated balloon. In order to
get some information about the stiffness of the myocardium one could
try to inflate the ventricle. The stiffer the myocardium is, the
higher pressure would be needed in order to increase the volume.
This relationship between the pressure and volume during the inactive
state is called the end diastolic pressure volume relationship
(\emph{EDPVR}).

@@

The ventricular pumping function depends on the loads that the
ventricle experience. If the amount of blood returning from the veins
into the heart increases, i.e increased preload, the end diastolic
volume increases and the ventricle contracts more forcefully in order
to synchronize the amount of ejected blood with the venous
return. This fundamental law, is called the \emph{Frank-Starling
  law}. The Frank-Starling law relates the EDPVR to the ESPVR. The law
states that the stroke volume, i.e the difference in end systolic and
end diastolic volume, increases in response to and increase in
preload. This allows the cardiac output to
be synchronized with the venous return and blood supply without
depending upon external regulation.

@@

% \begin{figure}[htbp]
%   \centering
%     \includegraphics[width=0.5\textwidth]{chapters/introduction/figures/Hill_muscle_model.pdf}
% \caption{The classical three-element Hill muscle model with one
%   contractile element and two non-linear springs, one arranged in
%   series and one parallel. }
% \label{fig:hill_muscle_model}
% \end{figure}

@@

\subsection{Cardiac Pathophysiology}
Pathophysiology is the study of the physiology of the diseased heart,
and we will briefly discuss some of the different cardiac diseases
encountered in this thesis.

\subsubsection{Heart failure}
Heart failure is a common term for all heart diseases in which the
heart in unable to supply the body with enough blood to meet its demand.
Consequently, patients suffering from heart failure has a reduced
cardiac output. It is common to distinguish between systolic and
diastolic heart failure.

@@

One type of heart failure can result from conduction block in the
heart. The action potential travels through these electrical highways
of high conducting cells. One of these highways known as the Bundle of
His splits into two parts, the right and left bundle. The right bundle
activates the right ventricle, while the left bundle activation the
left ventricle. For patients suffering from left bundle branch block,
there is ``road block'' on the highway along the left bundle, meaning
that the left ventricle is activated later than preferable, and the
right ventricle is typically activated prior to the left
ventricle. This can result in a dyssynchronous contraction, yielding a
reduced pumping effect. A promising treatment for this patients is
called Cardiac Resynchronization Therapy (CRT) \cite{cleland2005effect}.

@@

Experiments show that between 30-40\% of the selected patients do
not respond to CRT \cite{daubert20122012}, which is why there is much
research in  responder rate.

@@

\subsection{Summary of Cardiac Physiology}
The multiscale and multiphysics phenomena governening the mechanics of
the heart is complex. How the molecular dynamics should be coupled to
the electrophysiology at the cellular level, how the electrophysiology
should be coupled to the mechanics at the organ level and what type of
feedback mechanisms that should be included is far from fully
understood. In this thesis we limit ourself to the study of what is
happending on the organ level, and concentrates purely on the
mechanics.

>>>>

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.7\textwidth]{chapters/introduction/figures/heart_anatomy.pdf}
\caption{Schematic representation of the heart's anatomy. (Adapted and modified from Wapcaplet in Sodipodi).}
\label{fig:heart_anatomy}
\end{figure}

Deoxgenated blood is transported from the body trough the veins and to
the right atrium (RA). From the RA, blood flows to the right ventricle
(RV), before it is sent through the the pulmonary arteries(PA) and to the
lungs, where the bloods gets oxygenated. From the lungs, the blood
travels through the pulmonary veins to the left atrium (LA), and
finally to the left ventricle (LV), before is is sent out to the body
again through the aorta.

@@

Starting from the outside of the heart we have the \emph{pericardium},
the \emph{epicardium}, the \emph{myocardium} and the \emph{endocardium}. 
The endo- and epicardium are respectively the inner and outer layers of the
myocardium. The myocardium is made up by individual muscle cells, or
myocytes, which branch and join neighboring cells, and form a
complicated fibrous network which is often referred to as a 
\emph{functional syncytium} (meaning that the number of active muscle
fibers cannot be varied from beat to beat). Each myocyte is composed
of bundles of myofibriles, which again contains
\emph{myofilaments}. The myofilaments consist of a repeated pattern
of lines and bands which can be seen as a collection of individual
contractile units called \emph{sarcomeres}. A simple sketch of a
sarcomere is shown in Figure \ref{fig:sarcomere}. 
\begin{figure}[htbp]
  \centering
    \includegraphics[width=1.0\textwidth]{chapters/introduction/figures/Sarcomere.png}
\caption{Illustration of the sarcomere structure.}
\label{fig:sarcomere}
\end{figure}

The sarcomeres are composed thick and thin filaments that slides
back and forth when the muscle contracts or relaxes.
The thin filaments are made up by a protein called \emph{actin} while
the thick filaments are made up by a protein called \emph{myosin}.

@@

The heart beats in a cyclic manner, about one beat every second.
The contraction of the heart is triggered by an electrical signal called an action
potential originating from specialized cells in the right atrium. When an action
potential is triggered, calcium is released into the cells. These
calcium ions binds to a complex called troponin C located at the thin
filaments, which then exposes the actin to the myosin head. When
myosin binds to actin we say that a \emph{cross-bridge} is formed
between the thin and thick filaments. A \emph{power stroke} is
triggered after ATP hydrolysis, and the sarcomere shortens. 
The signal propagates through the myocardium and along
specialized electrical highways with high conducting cells.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.85\textwidth]{chapters/introduction/figures/cardiac_cycle.png}
\caption{Showing the relation between pressure and volume in the left
  ventricle during the cardiac cycle. To the left we show the a reduced version of
the classical Wiggers diagram with pressure and volume plotted against
time. To the right we show the pressure volume loop
with volume plotted on the $x-$axis and pressure on the $y-$axis. }
\label{fig:pv_loop}
\end{figure}

@@

The pressure and volume inside the ventricle can provide us with
information about the passive and active properties of the
myocardium. Imaging it was possible to shut down all contractile
units in the myocardium, and remove all loads (e.g set the pressure
to zero). This would be analogous to have a deflated balloon. In order to
get some information about the stiffness of the myocardium one could
try to inflate the ventricle. The stiffer the myocardium is, the
higher pressure would be needed in order to increase the volume.
This relationship between the pressure and volume during the inactive
state is called the end diastolic pressure volume relationship
(\emph{EDPVR}).

@@

The ventricular pumping function depends on the loads that the
ventricle experience. If the amount of blood returning from the veins
into the heart increases, i.e increased preload, the end diastolic
volume increases and the ventricle contracts more forcefully in order
to synchronize the amount of ejected blood with the venous
return. This fundamental law, is called the \emph{Frank-Starling
  law}. The Frank-Starling law relates the EDPVR to the ESPVR, and
states that the stroke volume, i.e the difference in end systolic and
end diastolic volume, increases in response to and increase in
preload. This allows the cardiac output to
be synchronized with the venous return and blood supply, without
depending upon external regulation.

@@

@@

\subsection{Cardiac Pathophysiology}
Pathophysiology is the study of the physiology of the diseased heart,
and we will briefly mention some of the different cardiac diseases
encountered in this thesis.

\subsubsection{Heart failure}
Heart failure is a common term for all heart diseases in which the
heart in unable to supply the body with enough blood to meet its demand.
% Consequently, patients suffering from heart failure has a reduced
% cardiac output. It is common to distinguish between systolic and
% diastolic heart failure.

@@

One type of heart failure can result from conduction block in the
heart. The action potential travels through these electrical highways
of high conducting cells. One of these highways known as the Bundle of
His splits into two parts, the right and left bundle. The right bundle
activates the right ventricle, while the left bundle activates the
left ventricle. For patients suffering from left bundle branch block,
there is ``road block'' on the highway along the left bundle, meaning
that the left ventricle is activated later than preferable, and the
right ventricle is activated prior to the left
ventricle. This can result in a dyssynchronous contraction, yielding a
reduced pumping effect. A promising treatment for this patients is
called Cardiac Resynchronization Therapy (CRT) \cite{cleland2005effect}.

@@

Experiments show that between 30-40\% of the selected patients do
not respond to CRT \cite{daubert20122012}, which is why much research
is put into increasing the responder rate.

@@

\subsection{Summary of Cardiac Physiology}
The multiscale and multiphysics phenomena governing the mechanics of
the heart is complex. How the molecular dynamics should be coupled to
the electrophysiology at the cellular level, how the electrophysiology
should be coupled to the mechanics at the organ level and what type of
feedback mechanisms that should be included is far from fully
understood. In this thesis we limit ourselves to the study of what is
happening on the organ level, and concentrates purely on the
mechanics.
